---
title: 'Week 1: Introduction'
output:
  pdf_document: default
  html_notebook: default
---


# Types of Data and Measurement Scales
- Measurement and data are not the same thing
  - Measurement is the process
  - Data is the output
  - Items/Units we measure

## Types of data
  - Quantitative data are data measures along a numerical       scale (continuous)
  - Qualitative Data Descriptions that fall into categories
    - Often referred to as discrete
    - Frequencies, proportions or rates (discrete)
    - Items/units we count
- Measurement & Measurement Scales
  - The numbers, or symbols, are assigned such that the         relationships amongst the numbers or symbols reflect     relationships in the attribute studied
  - measurements

## Type of Measurement Scales

###  Interval Scale
  - Numbers are assigned to observations such that differences between any two numbers correspond  to proportional differences in the underlying property studied, and there are equal intervals along the scale
  - Examples
    - Temperature measuresed in degrees Farenheit
    - Directional distance from a reference point
    - Calendar date
    - Height above sea level
  - differences in the measurement scale are the same as the underlying property
  - Permissible transformations include any linear transformation
  - Statistics such as the mean and standard deviation may be used
  - Most of the parametric statistical tests may be employed as long as their underlying assumptions are met
  - Interval scale values can be used to determine =, not equal, >, or <. Can also use sums and differences with meaning
  
###  Ratio Scale
  - numbers are assigned to observations such that an interval scale has been reached and there is a zero ppoint which corresponds to zero point which corresponds to a zero, null state or absence of the underlying property
  - Examples
    - Length
    - height
    - weight
    - distance
    - volume, weight
    - Cycle-time and time-to-repair
  - basically the same as interval scale but 0 means that there is nothing there. OR an absence of data
  - Good Transformations
    - Typically involve multiplying by a constant
  - Statistics Available are
    - The geometric mean
    - Coefficient of variation
    - Any statistical test as long as their underlying assumptions are met

## Ratio Scale vs Interval Scale
  
- Typically, when looking at a scale we think the set point is 0. Is 200 pounds twice as heavy as 100 pounds?
  - For ratio data the answer is yes
  - What if the scale was set at 50 to begin with? 
    - Then our scale is interval and the answer is no

### Absolute Scale

  - Numbers are assigned to observations such that the number directly correspond to the underlying property being studied.
  - Examples are
    - The number of defects
    - The number of scratches observed
    - The number of parts made
    - The number of safety accidents
    - The number of customer complaints
  - The underlying property is the same as the measurement count (or just say data)
  - Transformations
    - Some are permissible
    - Data on an absolute scale have some of the properties of ratio data
    - Mean, median and mode may be used
  - Various statistical procedures may be used depending on resolution
    - In many cases, standard parametric methods may be used. In other cases nonparametric methods must be employed

# Measurement as a Process: The 5 Big Aspects of Data

- We need to consider
  - Data costs money. We must make data both efficient and effective
  - How do we make data efficient and effective?
    - Make certain that the measurement process itself is effective
      - Capable
      - Acceptable

**Remember is a process, data is the output**

- Common Myths of Measurement
  - Is expensive
  - Has a digital readout
  - Uses radiation
  - Is made in Switzerland
  - Is only used by a Supervisor
  - Has a recent calibration sticker
  
- Measure measurement systems must demonstrate
  - Stability through time, or control
  - The ability to generate reliable or repeatable and reproducible measures
  - The ability to generate valide measures

**Reliability in measurement is a measure of the PRecision of the Device / methodf**

**Validity in metrology is a measure of the Accuracy of the device/methods **

- Precision vs Accuracy
  - Accuracy is the closeness of the desired target
  - Precision is the closeness of the predictions
    
## Tools for Understanding Data
  - There are many tools to help us understand data
    - Probability and statistics: to quanitfy and summarize data
    - Control charts: to determine if a process is stable
    - Experimental design: a;;s us to identify root causes of a problem so we can eliminate it or to identify and properly manipulate Special and Common Causes of Variability to improve optimize the process output
    
## Big 5 Aspects of Data

### Location
- Measures of location describe a middle or center point tendency of a distribution
  - Mean
  - Median
  - Mode
  
### Spread
  - Measures of dispersion reflect the variation or spread in a data set or distribution
  - Range
  - Standard Deviation
  - Variance
  
### Shape
  - Type of distribution samples
  - Skewness
    - degrees of departure from symmetry of a distribution
  - Kurtosis is concerned with the peakedness of the distribution
    - Leptokurtic is a high peak. Positive Kurtosis.
    - Mesokurtic is a normal peak
    - Platykurtic is a lower peak

### Time Sequence

### Measures of Relationship
- Correlation
- Association
  - Same as correlation but on of the variables is discrete

# Sampling Concepts

## Population and Samples
  - Population and Samples are not the same
  
### General Approaches to sampling
  - Nonrandom or judgement sampling
  - Random or Probability Sampling
    - Simple random Sampling
    - Systematic random sampling
    - Stratified random sampling
    - Cluster sampling

#### Nonrandom Sampling
  - Specimens or items are selected using personal judgement, reasoning, opinion or convenience

#### Random or probability sampling
  - All specimens or items have a probability of being included in the sample

#### Simple Random Sampling
  - Every possible sample of size n has an equal change of being selected

#### Systemic Random Sampling
  - Specimens or items are selected at an interval

#### Stratified Random Sampling
  - Specimens or items are divided into homogenous subsets or strata 
  
#### Cluster Sampling
  - Specimens or items are divided into groups that are homogenous between each other, but hetereogenous within

A **statistic** is a measure calculated from sample data that may be used to make inferences about a population
  - the average is a "statistic"
  - The range is another "statistic"
  - Ther are many more
  
**descriptice statistics**
  - Describe a characteristic of a sample
  - Frequently used to make inferences about population parameters
  - Represented by letters in english
**Popoulation Parameters**
- Describes a characteristic of the population
- Represented by Greek letters (with a few exceptions)

```{r install}
# install.packages("devtools")
# require(devtools)
# install_github("burrm/lolcat")
```

```{r load_package}
library(lolcat)
```

# Weekly Quiz: Practice Assessment 

Were working with thet airline dataset

1. Ticket Price is Continuous-Ratio
2. Number of Cancelled Fligths per Dat per 500 *
  - Discrete Nominal
3. Baggage Delivery Time to Carousel
  - Continuous Interval
4. Number of Lost Bages Per 100 Passenger *
  - Discrete-Ordinal-Count(Absolute)

# Assessment: Data and Measurement


    