---
title: "Airline_hmwrk"
author: "D. ODay"
date: '2022-06-23'
output: pdf_document
---
# Week 2

## Practive Assessment
```{r import}
airline <- read.delim("~/Documents/GitHub/school_cu/school_cu/methods for quality improvement/DTSA5704_DescribingData/data/airline.dat")

require(lolcat)
```
1. descriptive stat of mean: $\bar{X}$
1. descriptive stat of median: $\sim{X}$ (tilde over X) (Population is M)
3. descriptive stat of IQR: IQR
4. descriptive stat of Skewness: $g_{3}$
5. descriptive stat of kurtosis: $g_{4}$
6. parameter symbol variance: $\sigma^{2}$
7. parameter symbol skewness: $\gamma_{3}$
8. parameter symbol kurtosis: $\gamma_{4}$
```{r mean_variance}
mean(airline$Price)
var(airline$Price)
IQR(airline$Price)
rng = range(airline$Price)
rng[2] - rng[1]
skewness(airline$Price)
kurtosis(airline$Price)
```

# Assessment: Describing Data Graphically

1. The ideal number of lcass intervals to have i a grouped frequency distribution is 10.
2. 6
3. 6.649
4. 6.616,6.658
5. 0.24
6. 67.5
7. 10
8. 57.5, 112.5
9. postively skewed
10. 60, 100
11. 78.5
12. 265
13. has outliers
14. wrong: said 195, 305 and 0,8 should be 0, 0.2
15. the right section

```{r 2.2_3_4}
arr = c(6.649, 6.639, 6.629, 6.642, 6.624, 6.616, 6.649, 6.654, 6.649, 6.642, 6.649, 6.637, 6.649, 6.647, 6.644, 6.657, 6.654, 6.634, 6.644, 6.649)
df_arr = data.frame(arr)

#Run Chart
spc.run.chart(df_arr$arr)

#Ungrouped Freq Distr
frequency.dist.ungrouped(df_arr$arr)

#Ungrouped Polygon
frequency.polygon.ungrouped(df_arr$arr)

```

```{r 2.5_11}
viscosity <- read.csv("~/Documents/GitHub/school_cu/school_cu/methods for quality improvement/DTSA5704_DescribingData/data/viscosity.txt", sep="")

#Grouped Freq Distr
frequency.dist.grouped(viscosity$centistokes)

#Grouped Histogram
hist.grouped(viscosity$centistokes)

# 5 Num Sum
summary(viscosity$centistokes)

# Box and Whisker Plot
boxplot(viscosity$centistokes)
```

```{r 2.5_11}
yield <- read.csv("~/Documents/GitHub/school_cu/school_cu/methods for quality improvement/DTSA5704_DescribingData/data/yield.txt", sep="")

#Grouped Hist with Density
hist.grouped(yield$strength, freq=F)
lines(density(yield$strength))

plot(density(yield$strength))

hist.grouped(yield$strength)

```

#Describing Data Numerically
1. 6.643
2. 6.646
3. 6.649
4. negatively skewed
5. 0.01049
6. 0.00011
7. summary(), range(), frequency.dist.ungrouped()
8. 45
9. IQR, summary()
10. 78
11. quantile, summary()
12. -0.9062
13. negatively skewed
14. 0.6184
15. leptokurtic
16. continuous-ratio
17. 0.988
18. positive

```{r numeric_1_6}
mean(df_arr$arr)
median(df_arr$arr)
sample.mode(df_arr$arr)

sd(df_arr$arr)
var(df_arr$arr)
```

```{r numeric_7_11}
#returns skewness 
# summary.continuous(viscosity$centistokes)

# range(viscosity$centistokes)
IQR(viscosity$centistokes)

# frequency.dist.ungrouped(viscosity$centistokes)
summary(viscosity$centistokes)

#Percentile
quantile(x=viscosity$centistokes, probs=0.45)

```

```{r 12_15}
#skewness
round(skewness(yield$strength), 4)

#kurtosis
kurtosis(yield$strength)
```

```{r 16_18}
body <- read.csv("~/Documents/GitHub/school_cu/school_cu/methods for quality improvement/DTSA5704_DescribingData/data/body.txt", sep="")

cor(x=body$height, y=body$weight, method="pearson")
```