---
title: "Lets Talk About COVID"
output: pdf_document
indent: True
header-includes:
  - \setlength{\parindent}{2em}
  - \setlength{\parskip}{0em}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r tinytex_install}
# Can comment out if already have installed
tinytex::tlmgr_install("pdfcrop")
```

# Introduction

Four best friends hanging out in a living room. Three boxes of pizza, freshly 
delivered from Papa Johns. Two heated debates on who gets to play the infamous
Mario. And 1 Nintendo Switch loaded with one of our favorite games: Mario Party. 
We hit our die and finish our first minigame (I won of course). All of the sudden,
our cellphones begin to ring. Texts, news articles, and app feeds drown our 
phones with something we never heard before: COVID. Once we realized it wasn't 
insomnia alerting us about our cookies, we put down our phones and 
continued playing until one of us (sadly not me) became victorious. 
At that time, none of us knew it; that would be our last time 
together for months.  
COVID interrupted most of our lives for the past year and a half. Schools,
jobs, family events, and even concerts shifted to an online presence. 
Governments worked frantically to develop the best policies. Scientists worked
day and night to find a cure. And for some reason, toilet paper became a luxury
item in the United States. Regardless, COVID has changed lives on the global
scale. Throughout this data, we will review how COVID is trending and where we 
should focus so we can go back to a 'normal' life.

### Packages
First, we will need the **tidverse** and **lubridate** packages to carry out
our analysis.
```{r _load_packages}
library(tidyverse)
library(lubridate)
```

# Data
We will be using four csv files from the The New York Times Company* Github. 
The New York Times Company is an American mass media company that produces
a daily newspaper (*The New York Times*), located in New York City. This newspaper circulates both
domestically and internationally. The goal of the company is to deliver as much 
unbiased news information as possible. Using their data, we will
load the following datasets:

- time_series_covid19_confirmed_global.csv
    * Total amount of Covid Cases globally
- time_series_covid19_deaths_global.csv
    * Total amount of COVID-related deaths
- time_series_covid19_confirmed_US.csv
    * Total amount of COVID cases in the U.S.
- time_series_covid19_deaths_US.csv
    * Total amount of COVID-related deaths in the U.S.  

These datasets contains the number of daily cases from January 22, 2020 until
current date (In this case August 10, 2021).

```{r get_jhu_data}
url_in <- paste0("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/",
"master/csse_covid_19_data/csse_covid_19_time_series/")

filenames <- c("time_series_covid19_confirmed_global.csv", 
               "time_series_covid19_deaths_global.csv",
               "time_series_covid19_confirmed_US.csv", 
               "time_series_covid19_deaths_US.csv"
               )
urls <- str_c(url_in, filenames)
```

```{r import_data, message=FALSE}
global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])
```

# Data Cleaning
Lets take a quick look at the global_cases and global_deaths datasets.

```{r global_cases_view}
global_cases
```
```{r global_deaths_view}
global_deaths
```
Looking at the datasets, we need to do some cleaning. First, we will
put each date in one column called 'date'. This way, each date per country will represent
an observation. Also, we do not need the Latitude and Longitude for our
analysis, so we will drop them.

```{r tidy_global_data}
global_deaths <- global_deaths %>%
  pivot_longer( cols = - c('Province/State',
                           'Country/Region', Lat, Long),
                names_to = "date",
                values_to = "deaths") %>%
  select(-c(Lat,Long))

global_cases <- global_cases %>%
  pivot_longer( cols = - c('Province/State',
                           'Country/Region', Lat, Long),
                names_to = "date",
                values_to = "cases") %>%
  select(-c(Lat,Long))
```
Next, we need to combine global_deaths and global_cases so the cases and deaths
are in one dataset. Also, we see our date is currently a character data type. 
We will need to use the *lubridate* package to convert the date column to a date
object.

```{r tidy_join_global_data}
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))
```

```{r tidy_global_data_summary}
summary(global)
```
Looking at the data, we can see the cases are heavily, positively skewed;
ths could mean there are a lot of rows with 0. So, we want to filter those out.
After the filter, our minimum case is 1. Now lets view the see if the
maximum number of cases listed is an outlier.

```{r tidy_global_data_summary_filter}
global <- global %>% filter(cases > 0)
summary(global)
```
As we can see, there are multiple cases close to 36055002. So, we don't have to
worry about this data point being an outlier.

```{r tidy_global_filter_maax}
global %>% filter(cases > 35000000 )
```
Now we want to take a look at the US cases. When looking at the dataset we can
see some weird codes and data types. We need to pivot the dates, while keeping Admin2, 
Province/State, Country/Region and Lat/Long. We also need to convert *date* to a date 
object. We can do the same for us_deaths as it follows a similar format as
us_cases.

```{r tidy_us_cases_view}
us_cases
```

```{r tidy_us_cases_data}
us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
```

```{r tidy_us_deaths_data}
us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
```
We will join the two US datasets.
```{r tidy_us_join_data}
US <- us_cases %>%
full_join(us_deaths)
```

```{r tidy_us_datasets_final}
US
```
We need to combine the state and country_region variables of the global_dataset to 
create a key. This will allow some comparative analysis of the 
different countries. Also, we need to add the population of these countries to
the final dataset.

```{r tidy_global_country_region_data}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```
We will use this csv to get the population for the different countries.
``` {r tidy_population_csv_load}
uid_lookup_url <- paste0("https://raw.githubusercontent.com/CSSEGISandData/",
"COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv")

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```
Here we'll add the uid csv to global_dataset to add the population as a column.
```{r tidy_global_add_pop}
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, 
         Combined_Key)
global
```

# Visualization and Analysis

First, we'll create a data set that will have the number of cases and deaths by state.
Also, we will create the **deaths_per_mil** variable to use for comparative
analysis.

``` {r state_us_cases}
us_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
    Population = sum(Population)) %>%
    mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
    select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill,
    Population) %>%
    ungroup()
us_by_state
```
Next we can view the total amount of cases and deaths for the US. We can see
at the beginning of 2020 the first cases of COVID came to the US.

``` {r total_us_data}
us_totals <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

us_totals
```
Next, lets visualize the total amount of cases and deaths in the United States 
since the beginning of COVID. We can see the number of cases increases, substantially,
from the beginning to end of 2021. Since, there was a slight increase
until March; then the amount of cases have been stagnant.  During this time the
US  lifted their restrictions and the cases didn't drastically increase.
This could be due to the effects of the vaccine. More people were returning back
to normal lives without causing much effect on the population.

``` {r visual_us_cases}
us_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y = NULL)
```
Next, lets look at New York. This will help us get a glimpse of what is 
happening at the state level. At first glance, it seems that the number of COVID cases have
leveled off. Is this true?

``` {r visual_cases_new_york}
us_by_state %>%
  filter(Province_State == "New York") %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in New York", y = NULL)
```

# Analyzing the Data

As the numbers get larger, the total amount of cases loses its meaning. 
At first, fifty new COVID cases seemed like a lot. However, once we reached
500,000 cases, it's hard to tell the change on a daily bass. 
Lets add variables to represent the daily, new amount of cases and deaths.

``` {r add_lage_to_data}
us_by_state <- us_by_state %>%
  mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
us_totals <- us_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

us_by_state
```
Now, lets visualize the amount of new cases per day. We can see that there was a
decrease after March. Again, this is when many American citizens
began receiving the COVID vaccine. However, with the new Delta variant, we can
see a rise in July 2021. In fact, the amount is close to the 
highest daily amount since January 2021 - the peak of COVID cases.

```{r new_cases_data}
us_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US (Daily)", y = NULL)
```
Which states are handling COVID the best? Which are handling it the worst?
We'll see the 10 states 
with smallest/largest deaths per thousand. Looking at the first table,
we can see the states with the least deaths per thousand.
Most of these states are areas that are rural or tourist attractions.  Are Hawaii and the Virgin Islands
doing better due to the drop in tourist activity? On the other hand, Alaska and Utah are large states 
with a smaller population; citizens are more spread out and may not come in
contact as often.

``` {r best_states_covid_analyze}
us_state_totals <- us_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000* cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

us_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```
Compared to the states with the lower amounts of deaths per thousands, the 
states with the higher amount of cases have a larger population. Congested areas may have more
occurrences of contact with others; this may cause more cases/deaths. When 
comparing these states, we need to question how these cases/deaths recorded?
Also, we need to consider factors that may cause a 
difference that isn't recorded. One factor could be health care quality or 
citizens' access to health care.

``` {r worst_states_covid_analyze}
us_state_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```
## COVID by Political Party

Although COVID has taken another recent surge, the United States political
parties are still divided on what actions we should take. Republicans believe
the country should operate as normal and  the disease will either go away, or 
become something we adapt to. On the other hand, Democrats believe we should
take as much precaution as possible; also, Democrats have a stronger push for
national vaccination requirements. Lets take a look and see if the polarizing
views affect the political parties differently. 
First, we'll create vectors consisting of members of the political parties.
We're defining each state's political party as who they voted for in the 2020
presidential election. We discovered this from:  

<https://www.archives.gov/electoral-college/2020>. 
Guam, Puerto Rico, The Virgin Islands nor The Northern Mariana Islands voted in
the U.S. election; therefore, we will remove their cases. and not include
them in this part of the analysis.
```{r covid_party}
republican <-c("Alabama", "Alaska", "Arkansas", "Florida", "Idaho", "Indiana",
               "Iowa", "Kansas", "Kentucky", "Lousiana", "Mississippi",
               "Missouri", "Montana", "Nebraska", "North Carolina", 
               "North Dakota", "Ohio", "Oklahoma", "South Carolina", 
               "South Dakota", "Tennessee", "Texas", "Utah", "West Virginia",
               "Wyoming")
democrat <- c("Arizona", "California", "Colorado", "Connecticut", "Deleware",
              "District of Columbia", "Georgia", "Hawaii",
              "Illinois", "Maine", "Maryland", "Massachusetts", "Michigan",
              "Minnesota",  "Nevada", "New Hampshire", "New Jersey", 
              "New Mexico", "New York", "Oregon", "Pennsylvania", 
              "Rhode Island", "Vermont", "Virginia", "Wisonsin", "Washington")
```
Next, we need to create the 'party' column so we can label each state's
political party with how they voted in the 2020 presidential election.

```{r covid_party2}
us_by_party <- us_by_state %>% 
    filter(us_by_state$Province_State != "Guam" & 
           us_by_state$Province_State != "Puerto Rico" &
           us_by_state$Province_State != "Virgin Islands" &
           us_by_state$Province_State != "Northern Mariana Islands") %>%
  group_by(Province_State) %>%
  mutate(party = ifelse(Province_State %in% republican,'R','D')) %>%
  group_by(party, date) %>%
  summarise(cases = sum(cases),
            deaths = sum(deaths),
            population = sum(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
us_by_party
```
First, let's view how COVID surged in each party. We can see the democratic party
suffered from COVID a bit earlier than the republican party; however, it seems 
both became stagnant around January 2021.
```{r covid_party_plot_cases}
us_by_party %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases_per_thou)) +
  geom_line(aes(color = party)) +
  geom_point(aes(color = party)) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US by Political Party (Cases Over Time)", y = NULL) +
  scale_color_manual(breaks = c("D", "R"), values=c("blue", "red"))
```
Similar to the number of cases, states in the democratic party suffered from
more deaths than the republican party early on; however, the number of cases seemed
to even out towards the end.

```{r covid_party_plot_deaths}
us_by_party %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = deaths_per_thou)) +
  geom_line(aes(color = party)) +
  geom_point(aes(color = party)) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US by Political Party (Deaths Over Time)", 
       y = NULL) +
  scale_color_manual(breaks = c("D", "R"), values=c("blue", "red"))
```
As we know, a look at the overall amount of cases is not a good measurement of 
how COVID is affecting the country. Next we're looking at the number of new
cases each day. As we can see, after July of 2020, the amount of new COVID cases
seem to be the same. 

```{r covid_party_plot1}
us_by_party %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = party)) +
  geom_point(aes(color = party)) +
    scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US by Political Party (New Cases)", y = NULL) +
  scale_color_manual(breaks = c("D", "R"), values=c("blue", "red"))
```
Similar to the number of new cases, the number of daily new deaths has been
identical between political parties (other than between March 2020 - July 2020).

```{r covid_party_plot2}
us_by_party %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = new_deaths)) +
  geom_line(aes(color = party)) +
  geom_point(aes(color = party)) +
    scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US by Political Party (Death Rate)", y = NULL) +
  scale_color_manual(breaks = c("D", "R"), values=c("blue", "red"))
```
How many of the COVID cases resulted in deaths? Does the state's political party
make a difference? First, lets create a variable called "Death Rate" which will
be the number of deaths divided by the number of cases. As we can see, there seems to be an
outlier around March 2020. Let's take a closer look at these cases.

```{r us_by_party_death_rate}
us_by_party <- us_by_party %>%
  mutate(death_rate = 100* (deaths / cases))
us_by_party_rate_plot <- us_by_party %>%
  filter(cases > 0 & death_rate != Inf) %>%
  ggplot(aes(x = date, y = death_rate)) +
  geom_line(aes(color = party)) +
  geom_point(aes(color = party)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US by Political Party (Death Rate)", 
       y="Death Rate %") +
  scale_color_manual(breaks = c("D", "R"), values=c("blue", "red"))
us_by_party_rate_plot
```
These dates are from the beginning of COVID when there weren't many cases.
 Let's recreate the plot while ignoring these outliers.
```{r find_death_rate_outlier}
us_by_party %>% filter(death_rate > 10 & death_rate != Inf) %>%
  select(c("party", "date", "cases", "deaths", "death_rate"))
```
Looking at the graph again, it seems the death rate has become rather
consistent. Since there's a large amount of total cases, maybe it would help to
define death rate as the number of new_deaths divided by the number of 
new_cases.
```{r death_rate_plot_no_outlier}
us_by_party_rate_plot + ylim(0, 12)
```
Disregarding the three outliers in March of 2020, we can see the death rate of
new cases has fluctuated between 0 and 5%, for the most part. In fact, this past
month (July 2021) has had a period of lowest death rates since COVID began. 
Similar to the other cases between COVID data, there isn't much of a difference
between political parties. Maybe the United States push for vaccines have
helped? 

```{r plot_death_rate_party_lag_table}
us_by_party <- us_by_party %>%
  mutate(death_rate = 100* (new_deaths / new_cases))
us_by_party_rate_plot <- us_by_party %>%
  filter(cases > 0 & death_rate != Inf) %>%
  ggplot(aes(x = date, y = death_rate)) +
  geom_line(aes(color = party)) +
  geom_point(aes(color = party)) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US by Political Party (Daily Death Rate)", 
       y="Death Rate %") +
  scale_color_manual(breaks = c("D", "R"), values=c("blue", "red"))

us_by_party_rate_plot + ylim(0,15)

```

# Modeling Data
After analyzing the data, we would like to build linear model to help predict
the number of cases in the future. In other words, we would like to predict 
the number of deaths per thousand, given the number of cases.
Looking at the model, it is telling us that we can get the number of deaths per
thousand if we subtract .018 from .016 * cases per thousand.

``` {r linear_model_summary}
  mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)

```
Here we can compare the actual predictions of deaths compared to the actual
values. We can see, other than Alaska, Connecticut, and the District of 
Columbia, our predictions are close.
```{r linear_model_pred}
us_tot_w_pred <- us_state_totals %>% mutate(pred = predict(mod))
us_tot_w_pred %>% select(c('Province_State', 'cases_per_thou',
                       'deaths_per_thou', 'pred')) %>% head(10)
```
Let's plot these predictions with our real data. We can see our prediction
(in red) follows the same trend as the real COVID data (in blue). The model
makes an exact prediction for some and it's largely off for some. It would be
great to look further to see which factors are causing this issues.

``` {r plot_predictions}
us_tot_w_pred %>% ggplot() + geom_point(aes(x=cases_per_thou, y=deaths_per_thou),
                                        color = "blue")+
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```

# Conclusion

  When viewing this data, it is important to consider potential bias. First,
using different variables could lead to different results. For example, 
implementing population density could have made a difference in our predictions.
The closer in contact people are with COVID, the more likely they are
to develop symptoms. Also, I chose not include the outliers in the Death Rate
plots. Removing them helped us view the difference in the smaller data points. If
I wouldn't have removed the outliers, the view of the graph would have made it
seem as if the death_rates didn't change. Third, I have my own opinions bout
COVID; therefore, I made choices to review different factors than others might have. For
example, I chose to look at the differences between political parties when I
could have viewed the differences between regions; or, I could have studied the
data on a global scale.  
Overall, it seems COVID is not disappearing any time soon. In fact, it seems
that we're having another surge. However, just because we are still
having COVID cases, doesn't mean we are experiencing deaths as much as we have in
the past. This could be because of government policy, citizen interactions,
vaccine development or other potential factors. Although the U.S. political parties
have opposing viewpoints, these opinions don't create different experiences with COVID.
Regardless, we should keep developing strategies so we can
all go back to our normal lives, while being safe. After all, I still haven't 
gotten my rematch of Mario Party!

```{r source_info}
utils::sessionInfo()
```