---
title: "NYPD Data"
output: pdf_document
---

```{r tinytex_install}
#Can comment out if already installed
tinytex::tlmgr_install("pdfcrop")
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



## Packages Needed
- Tidyverse
- Lubridate
- DT
```{r import_packages, message= FALSE}
library(tidyverse)
library(lubridate)
library(DT)
```


## Importing the data

First, I'll import the data from <https://catalog.data.gov/dataset/nypd-shooting-incident-data-year-to-date> . This data represents information about every shooting incident in New York City since 2006.

```{r data_import}
url_nypd <- paste0("https://data.cityofnewyork.us/api/views/833y-fsy8/",
              "rows.csv?accessType=DOWNLOAD")

nypd_shootings <- read_csv(url_nypd)

```

## Tidying Data

Looking at the column details, I can see some columns are not the correct 
variable types. Therefore, I will make the following changes

- *Occur_Date* is listed as a string/character type
  - This needs to change to a date column using the **lubridate** package
- The following variables will need to be changed to a factor type because they
are categorical
  - *BORO*
  - *JURISDICTION_CODE*
  - *PERP_AGE_GROUP*
  - *PERP_SEX*
  - *PERP_RACE*
  - *VIC_AGE_GROUP*
  - *VIC_SEX*
  - *VIC_RACE*
- I'm also removing a few variables that I don't feel have as much impact to the 
analysis. INCIDENT_KEY would be important if we were joining multiple datasets.
In this case, we aren't; therefore, I am removing it along with the geographical
data. **LOCATION_DESC** can be very useful; however, at first glance it seems as
if there is a lot of missing data. First we'll take a look at the missing
amount.

```{r loc_desc}
sum(is.na(nypd_shootings$LOCATION_DESC)) / nrow(nypd_shootings)
```

Because over half of the data is missing, we will remove **LOCATION_DESC** as
well.

```{r tidy_col_types}
factor_cols <- c("BORO", "JURISDICTION_CODE", "PERP_AGE_GROUP", "PERP_SEX",
                 "PERP_RACE", "VIC_AGE_GROUP", "VIC_SEX", "VIC_RACE")

nypd_shootings <- nypd_shootings %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE),
                                            across(.cols = all_of(factor_cols), 
                                                   as.factor)) %>%
  select(-c(INCIDENT_KEY, X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat,
            LOCATION_DESC))
```
Viewing the summary, we can see that about of a third of the PERP_AGE_GROUP,
PERP_SEX, AND PER_RACE are missing. Thus, I will drop all rows that are missing
data in these columns.If we had access to more data,  I could probably fill the 
missing data using various methods. Also, JURISDICTION_CODE only has two 
observations where the data is missing, I will fill them with a random number
between 0 and 2.

```{r data_summary}
summary(nypd_shootings)

```


```{r missing_vals}
nypd_shootings <- nypd_shootings %>%
  mutate(JURISDICTION_CODE = replace(JURISDICTION_CODE, is.na(JURISDICTION_CODE)
                                     , sample(0:2, 1))) %>%
  drop_na(PERP_AGE_GROUP, PERP_SEX, PERP_RACE)
sprintf("The number of missing values is: %i", sum(is.na(nypd_shootings)))

summary(nypd_shootings)
```
From the summarize table, we can see that there are three 'Other' variables.
As we can see below, these age groups seem as if they're typos. Therefore, we 
will change the values to unknown.

```{r perp_age_other}
nypd_shootings %>% filter(nypd_shootings$PERP_AGE_GROUP != "18-24" &
                          nypd_shootings$PERP_AGE_GROUP != "25-44" &
                          nypd_shootings$PERP_AGE_GROUP != "UNKNOWN" &
                          nypd_shootings$PERP_AGE_GROUP != "<18" &
                          nypd_shootings$PERP_AGE_GROUP != "45-64" &
                          nypd_shootings$PERP_AGE_GROUP != "65+"
)

nypd_shootings["PERP_AGE_GROUP"][nypd_shootings["PERP_AGE_GROUP"] == "1020" |
                              nypd_shootings["PERP_AGE_GROUP"] == "940" |
                              nypd_shootings["PERP_AGE_GROUP"] == "224"] <-
  "UNKNOWN"
                                                            
```
## Visualization and Analyzation
First, lets view the summary of the data
```{r first_summary}
summary(nypd_shootings)
```

First, we can see that about 19% of the all of the shootings were labeled as
murders.

```{r statistical_murder_intro}
ggplot(nypd_shootings, aes(x=STATISTICAL_MURDER_FLAG)) +
  geom_bar() +
  labs(title = "Cases Labeled as Murders", x = "Murder") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
  
```
Maybe the amount of shooting incidents differ between different boroughs? It
seems as if there are more shootings between the Bronx and Brooklyn
compared to others. However, it seems the percentage of these shootings that
are labeled as murders is consistent across all.

```{r shootings_boro}
ggplot(nypd_shootings, aes(x=BORO)) +
  geom_bar() +
  labs(title = "Shootings per Borough ", x = "Borough") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
  
```



```{r statistical_murder_boro}

nypd_shootings %>% group_by(BORO) %>% summarise(
  total_shootings = n(),
  statistical_murder = sum(STATISTICAL_MURDER_FLAG == TRUE),
  percentage = statistical_murder / total_shootings) %>%
  arrange(desc(percentage)) %>%
  rename("Cases" = total_shootings, "Murder Label" = statistical_murder,
         "%" = percentage)

```

Next, differing age groups may have different experiences within the city.
Therefore, there may be different reasons for shootings. We can tell by the 
graphs below there are more shooting incidents between perpetrators of 18-44 
years; however, perpetrators aged 45 years or older had a higher proportion of
cases being labeled as a murder.

```{r shootings_perp_age}
ggplot(nypd_shootings, aes(x=PERP_AGE_GROUP)) +
  geom_bar() +
  labs(title = "Shootings by Age Perpetrator Group ", 
       x = "Perpetrator Age Group") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
  
```

```{r statistical_murder_perp_age}
nypd_shootings %>% group_by(PERP_AGE_GROUP) %>% summarise(
  total_shootings = n(),
  statistical_murder = sum(STATISTICAL_MURDER_FLAG == TRUE),
  percentage = statistical_murder / total_shootings) %>%
  ggplot(aes(x = PERP_AGE_GROUP, y = percentage)) +
  geom_col() +
  labs(title = "Murder % by Perpetrator Age Group ", 
       x = "Perpetrator Age Group", y = "%") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))

```
As stated before, different age groups have different experiences within New
York City. Similar to the perpetrator, there are a higher number of cases in
which the victim was aged between 18-44 while victims aged 45+ years had a
higher proportion of their cases labeled as a murder.

```{r shootings_vic_age}
ggplot(nypd_shootings, aes(x=VIC_AGE_GROUP)) +
  geom_bar() +
  labs(title = "Shootings by Victim Age Group ", x = "Victim Age Group") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
  
```

```{r statistical_murder_vic_age}
nypd_shootings %>% group_by(VIC_AGE_GROUP) %>% summarise(
  total_shootings = n(),
  statistical_murder = sum(STATISTICAL_MURDER_FLAG == TRUE),
  percentage = statistical_murder / total_shootings) %>%
  ggplot(aes(x = VIC_AGE_GROUP, y = percentage)) +
  geom_col() +
  labs(title = "Murder % by Victim Age Group ", x = "Victim Age Group", y = "%") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))

```
Next, viewing the differences between the quantity of cases among the age groups
compared to the differences between murder proportion made me curious to view
the differences between the perpetrator/victims sex. For both the victim and
perpetrator, males were involved in more shootings compared to females; however,
a higher percentage of female cases were considered murders compared to males.

```{r shootings_perp_sex}
ggplot(nypd_shootings, aes(x=PERP_SEX)) +
  geom_bar() +
  labs(title = "Shootings by Perpetrator Sex ", x = "Perpetrator Sex") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
  
```

```{r statistical_murder_perp_sex}
nypd_shootings %>% group_by(PERP_SEX) %>% summarise(
  total_shootings = n(),
  statistical_murder = sum(STATISTICAL_MURDER_FLAG == TRUE),
  percentage = statistical_murder / total_shootings) %>%
  ggplot(aes(x = PERP_SEX, y = percentage)) +
  geom_col() +
  labs(title = "Murder % by Perpetrator Sex ", x = "Perpetraor Sex", y = "%") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))

```

```{r shootings_victim_sex}
#Done
ggplot(nypd_shootings, aes(x=VIC_SEX)) +
  geom_bar() +
  labs(title = "Shootings by Victim Sex ", x = "Victim Sex") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
  
```

```{r statistical_murder_victim_sex}
nypd_shootings %>% group_by(VIC_SEX) %>% summarise(
  total_shootings = n(),
  statistical_murder = sum(STATISTICAL_MURDER_FLAG == TRUE),
  percentage = statistical_murder / total_shootings) %>%
  ggplot(aes(x = VIC_SEX, y = percentage)) +
  geom_col() +
  labs(title = "Murder % by Victim Sex ", x = "Victim Sex", y = "%") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))

```
Lastly, many of the shootings involved a Black American perpetrator and/or
victim. However, a higher percentage of cases involving White Americans were
labeled as murders.

```{r shootings_perp_race}
ggplot(nypd_shootings, aes(x=PERP_RACE, fill = PERP_RACE)) +
  geom_bar() +
  labs(title = "Shootings by Perpetraor Race", x = "Perpetrator Race") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5)) +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  guides(fill=guide_legend(title="Perpetrator Race"))
  
```

```{r statistical_murder_perp_race}
nypd_shootings %>% group_by(PERP_RACE) %>% summarise(
  total_shootings = n(),
  statistical_murder = sum(STATISTICAL_MURDER_FLAG == TRUE),
  percentage = statistical_murder / total_shootings) %>%
  ggplot(aes(x = PERP_RACE, y = percentage, fill = PERP_RACE)) +
  geom_col() +
  labs(title = "Murder % by Perpetrator Race ", x = "Perpetraor Race", y = "%") +
  theme(axis.title.x = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  guides(fill=guide_legend(title="Perpetrator Race"))

```

```{r shootings_victim_race}
ggplot(nypd_shootings, aes(x=VIC_RACE, fill = VIC_RACE)) +
  geom_bar() +
  labs(title = "Shootings by Victim Race", x = "Victim Race") +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  guides(fill=guide_legend(title="Victim Race"))
  
```

```{r statistical_murder_victim_race}
nypd_shootings %>% group_by(PERP_RACE) %>% summarise(
  total_shootings = n(),
  statistical_murder = sum(STATISTICAL_MURDER_FLAG == TRUE),
  percentage = statistical_murder / total_shootings) %>%
  ggplot(aes(x = PERP_RACE, y = percentage, fill = PERP_RACE)) +
  geom_col() +
  labs(title = "Murder % by Victim Race ", x = "Victim Race", y = "%") +
  theme(axis.title.x = element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  guides(fill=guide_legend(title="Victim Race"))

```
The difference in murder percentages compared to the counts of incidents can be
due to cultural differences. For example, there are many males that love to
hunt. Hunting is physically demanding; therefore, many hunters are younger to 
middle-aged. Accidents that occur during hunting can be considered a shooting
but wouldn't be labeled as a murder. This difference would cause the number
of shooting incidents for males aged between 18 and 44 to increase. Since
hunting is not as common among females, less of their cases would be labeled as
shooting incidents as well. In summary, the demographic that uses guns more
may have more shooting incidents, not labeled as a murder, because 
they are handling them more.

Next, boroughs have different lifestyles due to location and differing financial
situations. However, we can see that murder cases involving 18-44 year old
citizens is consistently common across all boroughs.

```{r statistical_murder_age_vs_boro}
murder_df <- nypd_shootings %>% filter(STATISTICAL_MURDER_FLAG == TRUE)
murder_df %>%group_by(BORO, VIC_AGE_GROUP) %>%
  ggplot(aes(x = BORO, fill = VIC_AGE_GROUP)) +
  geom_bar() +
  labs(title = "Murder Cases by Victim Age Group per Borough", 
       x = "Borough", y = "Cases Labeled as Murder") +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))+
  guides(fill=guide_legend(title="Victim Age Group"))

```
Many citizens are involved in activities and cliques with people of similar age.
Does this cause murders where the perpetrator and the victim are the same age?
As shown below, although it's not a staggering difference, slightly over half of 
the murder cases involve situations where the perpetrator and the victim are of
different age groups.

```{r statistical_murder_same_age}
murder_df %>% mutate(same_age = ifelse(as.character(PERP_AGE_GROUP) == 
                                              as.character(VIC_AGE_GROUP),
                                            TRUE, FALSE)) %>%
  group_by(same_age) %>%
  ggplot(aes(x=same_age)) +
  geom_bar() +
  labs(title = "Murders on People Same Age", x = "Same Age") +
  scale_x_discrete(labels = c("TRUE" = "Same", "FALSE" = "Different")) +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
  
```
Contrary to age, murders are common among victims of the same race. This could
be due to cultural similarities.

```{r statistical_murder_same_race}
murder_df %>% mutate(same_race = ifelse(as.character(PERP_RACE) == 
                                              as.character(VIC_RACE),
                                            TRUE, FALSE)) %>%
  group_by(same_race) %>%
  ggplot(aes(x=same_race)) +
  geom_bar() +
  labs(title = "Murders on People Same Race", x = "Same Race") +
  scale_x_discrete(labels = c("TRUE" = "Same", "FALSE" = "Different")) +
  theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
```
Do murder incidents occur at different times during the day? month? year?
Compared to murder cases there is no disparity between a shooting incident 
during the time of day, month or year. First, all cases are more frequent during
later hours into the early morning. Also, more incidents occur during
the summer months compared to other seasons. This could be because of the
increase in weather temperature; more people would like to go outside with 
friends and family causing more interpersonal contact. I doubt anyone wants to
be outside during the winter; New York City winters can be quite brutal! Lastly,
there seems to be no relationship between the day of the month and shooting
incidents, whether they are labeled a murder or not.

```{r statistical_murder_time}
non_murder_df <- nypd_shootings %>% filter(STATISTICAL_MURDER_FLAG == FALSE)

nypd_shootings <- nypd_shootings %>% mutate(day = format(OCCUR_DATE, "%d"),
                          month = format(OCCUR_DATE, "%m"),
                          year = format(OCCUR_DATE, "%y"))

ggplot(nypd_shootings, aes(x=OCCUR_TIME, color = STATISTICAL_MURDER_FLAG)) +
  geom_histogram(binwidth = 3600, fill="white") + #Every hour
  guides(color=guide_legend(title="Murder Label?"))+
    theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5)) +
  labs(title = 'Murder vs Nonmurder During Day', x = 'Time of Day (hours)')
  
  
```


```{r statistical_murder_days_of_month}
nypd_shootings %>% group_by(month, STATISTICAL_MURDER_FLAG) %>%
  summarise(incidents = n()) %>%
  ggplot(aes(x=month, y = incidents)) +
  geom_point(aes(color=STATISTICAL_MURDER_FLAG)) +
  labs(title = "Murders vs Non-Murders Monthly", x= "Month", y="Incidents") +
  guides(color=guide_legend(title="Murder Label"))+
    theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5)) +
  scale_x_discrete(labels = c("Jan", "Feb", "Mar", "Apr", "May", "June", "July",
                              "August", "Sept", "Oct", "Nov", "Dec"))
```

```{r statistical_murder_days}
nypd_shootings %>% group_by(day, STATISTICAL_MURDER_FLAG) %>%
  summarise(incidents = n()) %>%
  ggplot(aes(x=day, y = incidents)) +
  geom_point(aes(color=STATISTICAL_MURDER_FLAG)) +
  labs(title = "Murders vs Non-Murders Day of Month", x= "Day", y="Incidents") +
  guides(color=guide_legend(title="Murder Label"))+
    theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
```
In a perfect world, everyone would love to live in a city where there is no
violence. Although that is not a reality today, New York city has
significantly reduced the number of shooting incidents and statistical murders 
since 2006. That being said, the current policies and cultural lifestyle
are improving the quality of life.

```{r statistical_murder_yearly}
#look at the times of the year the murders occur by month vs non murder


nypd_shootings %>% group_by(year, STATISTICAL_MURDER_FLAG) %>%
  summarise(incidents = n()) %>%
  ggplot(aes(x=year, y = incidents)) +
  geom_point(aes(color=STATISTICAL_MURDER_FLAG)) +
  labs(title = "Murders vs Non-Murders 2006-2020", x= "Year", y="Incidents") +
  guides(color=guide_legend(title="Murder Label"))+
    theme(axis.title.x = element_text(margin = 
                                      margin(t = 10)),
        plot.title = element_text(hjust = 0.5))
  
```

# Model Building
After our analysis, we would like to create a model to predict whether a
shooting would be considered a murder or not. First, we would like to drop a 
few variables that we don't believe are important or will be redundant in our
model.

```{r model_drop_vars}
nypd_shootings <- nypd_shootings %>% select(
  -c("OCCUR_DATE", "PRECINCT", "JURISDICTION_CODE", "day", "year"))
nypd_shootings$month <- as.factor(nypd_shootings$month)
```

Now, we can finally build our model!
```{r model_build}
logit_1 <- glm(STATISTICAL_MURDER_FLAG ~., family = binomial, 
               data=nypd_shootings)

summary(logit_1)
```

As we can see from the model, the variables with the coefficients that are
positive (all perpetrator age types except 'unknown', an unknown perpetrator
sex, or the incident occurring in April) increase the probability that the
incident is a murder case. On the other hand, a negative coefficient (the month
of July or November, or a male perpetrator ) decreases the likelihood of the
shooting being a murder case. It's quite surprising that the summer months
negatively impact if a case was murder or not!

# Conclusion
Statistical murders follow the same patterns and only account for 20% 
of all shootings. First, most cases involve males between the ages of 
18 and 44. Although Brooklyn and the Bronx have a higher number of incidents in
their boroughs, the percentage of those shootings that are murders is similar
to the other areas. On the other hand, cases in which the perpetrator and/or
victim has an age of 65+ are more likely to be considered a murder case. The
same can be said about cases involving a female and/or white perpetrator/victim.
Shootings involving the same race are likely to be considered murders while
ones involving different different age groups are not. Lastly, murders do not
differentiate from regular shootings when discussing chronological data. 
Both regular shootings and murders will increase during the night time and
during summer months; this could be due to warmer weather, or just a little more 
free time.

It is possible that there is bias contained within this report. First, we
only have a few variables to look at. Income disparities, population density, 
etc. are important variables to consider when looking at this data. 
Also, most of our data was categorical; this makes it difficult to compare
relationships of New York City. Maybe a higher population density would lead 
to a higher proportion of shootings being  a murder? More data could help us 
look at other factors that contribute to the differences between shooting types.

Third, the variables in this dataset have high multicollinearity. In other 
words, many of the variables aren't independent of each other; therefore, 
increasing or decreasing one variable in the model may cause another to increase
or decrease unintentionally. This causes major statistical errors when 
attempting to make predictions. 

Fourth, I wanted to avoid looking into specific variables due to ethical issues. 
Race was a variable I used in the model; however, I didn't want to investigate 
it much due to different issues an African American my face compared to a 
Caucasian or vice versa. Investigating racial differences can be tricky when 
working with data.

Lastly, I have my own personal biases. I may have made some choices on how to 
look at certain parts of the data, subconsciously. For example, I chose to focus
on whether a shooting was considered a murder. However, I could of focused on 
the disparities between different boroughs. These decisions could lead to 
different interpretations of the data.

In conclusion, New York City has a long history of gun violence. For the last 
14 years they have done a great job of reducing the incidents. Lets focus on 
reducing the violence more to make New York great for everyone.

```{r session_info()}
utils::sessionInfo()
```